---
title: 'Data visualization of built-in earthquake data `quakes`'
author: 
- name:Monika Baloda
  email:mbalo005@ucr.edu
output:
  html_document:
    toc: yes
---


This Rmarkdown does the followings: 

1. Visualize using following plots:
    i)density histogram ii) box-plot  iii) empirical cdf  iv) Q-Q plot 

    
2. Outlier detection 

3. Writing code to calculate Minimum variance unbiased estimates for Normal distribution. 

4.  Visualize your model fitness by plotting standard normal curve over sample. 

5. Multiple scatterplots in one plot in 2x2 layout.

6. Single scatterplot 

7. Plotting on Map


- This work uses **R Base Graphics** to generate the figures, not `ggplot2`.


**0. Load necessary packages**
```{r, collapse=T}
library(datasets) # for the `quakes` data set

## You need to install the maps package first, then you can comment out the following line.
# install.packages("maps")

library(maps) # for map visualization
```

The `quakes` data set give the locations of 1000 seismic events of MB > 4.0. The events occurred in a cube near Fiji since 1964.

```{r}
library(datasets) # we have already loaded the package but we can do it again
?quakes 
```

Let's get a quick glimpse of the dataset.
```{r, collapse=T}
class(quakes)
head(quakes, n=5) #print first 5 rows of quakes
dim(quakes)  # dimension of the table
names(quakes)  # list the variables in quakes
str(quakes)  # list the structures in quakes
```

***1.  Visualizations of the earthquake magnitudes***

Suppose we are interested in studying the distribution of the magnitude `mag` variable.  Following R code makes a single figure with the following four subfigures in a 2-by-2 layout.  
i)  subfigure #1: plot a density histogram of the earthquake magnitudes, and then plot the estimated probability density curve in red color in the same plot
ii) subfigure #2: plot a horizontal box plot of the earthquake magnitudes
iii)subfigure #3: plot the empirical cdf of the earthquake magnitudes
iv) subfigure #4: make a Q-Q plot to compare the observed earthquake magnitudes distribution with the theoretical Normal distribution. Add a *thick* qqline in blue color.  

**ANSWER**
```{r, collapse=T}
# part a)
par(mfrow=c(2,2))
hist(quakes$mag, freq=F, xlab="Earthquake Magnitude") # code for histogram
lines(density(quakes$mag), col="red", lwd=3) # superimposed edf

# part b)
boxplot(quakes$mag, horizontal=T, col=2, border="blue",
xlab="Earthquake Magnitude", main="Boxplot")

# part c)
plot.ecdf(quakes$mag, verticals=T, pch="",
xlab="Earthquake mag",
main="ECDF of Earthquake Mag") # empirical CDF plot

# part d)
qqnorm(quakes$mag, main = "Q-Q Plot of Earthquake Magnitudes")
qqline(quakes$mag, col="blue", lty=5, lwd=3)
```

***2. Outlier detection***

Following code prints out the indexes of the outliers and then prints out the outlier observations.  

```{r, collapse=T}
which(quakes$mag %in% boxplot(quakes$mag)$out)
```

How many outliers have you found?

answer: There are 7 outliers

**ANSWER**
```{r, collapse=T}
outliersValue <- boxplot(quakes$mag)$out
print(outliersValue)
```


***3. Minimum variance unbiased estimates for Normal distribution***

Suppose we assume that the earthquake magnitude variable follows a Normal distribution and our 1000 earthquake magnitude observations are independent and identically distributed (iid). 

That is, $X_i \sim N(\mu, \sigma^2)$, where $X_i$ is the magnitude of the $i$-th earthquake observation and $i=1,2,\cdots,1000$.  

The **minimum variance unbiased estimators (MVUE)** for $\mu$ and $\sigma^2$ are:
$$
\widehat{\mu}=\overline{X}=\frac{1}{n}\sum_{i=1}^n X_i \\
\widehat{\sigma^2}=S^2=\frac{1}{n-1}\sum_{i=1}^n (X_i - \overline{X})^2
$$

Using existing R functions to calculate the estimates of $\widehat{\mu}$ and $\widehat{\sigma^2}$.

**ANSWER**
```{r, collapse=T}
mean(quakes$mag)
(sd(quakes$mag))^2
```

Creating our code to calculate the estimate of minimum variance unbiased estimators (MVUE) i.e. $\widehat{\mu}$ and $\widehat{\sigma^2}$.

**ANSWER**
```{r, collapse=T}
x <- sum(quakes$mag)*(1/length(quakes$mag))
s2 <- (1/((length(quakes$mag)-1)))*sum(((quakes$mag - x)^2))

```

***4 Visualize your model fitness***

To visualize the model fitness, we can add the estimated Normal distribution curve to the histogram plot you have generated in part 1).

Our code generates the following figure.  

- make a density histogram first, set `xlim` from 0 to 8.
- plot the empirical density curve in red color on the same figure.
- plot the estimated Normal distribution curve (that is, $X \sim N(\widehat{\mu}, \widehat{\sigma^2})$) as a blue dashed line on the same figure.

```{r, collapse=T}
hist(quakes$mag, freq = FALSE, main = "Histogram of Earthquake Magnitudes", xlab = "Earthquake Magnitudes", xlim = c(0,8))
lines(density(quakes$mag), col="red")
z <- seq(0, 8, length=1000)
prob.z <- dnorm(z, mean=x, sd = sd(quakes$mag))
lines(z, prob.z, lwd=2, lty=2, col="blue")
```
*Questions* How does your estimated normal curve differ from the empirical density curve? :Do you think the earthquake magnitude observations follow a Normal distribution? Does this result consist with the Q-Q plot you generated in part 1).

**ANSWER**
The red curve and the blue line are quite similar, so the earthquake magnitude observations may follow a Normal distribution and it consist with the Q-Q plot.

***5. Pairwise scatterplots***
This figure tells us about the relations of every two variables in the dataset
```{r}
# require(graphics) # load the graphics library
library(graphics)
pairs(quakes, main = "Fiji Earthquakes, N = 1000", cex.main=1.2, pch=".")
```
*Question*:  Do you think the `mag` variable and `stations` variables are positively correlated? Explain your answer.
*Answer*: Yes, I think they are positively correlated because in the figures in the right bottom, the plots scatters in a left-bottom and top-right pattern.



***6. Single scatterplot***

Following R code reproduces the scatterplot for the `mag` and `stations` variables.  

- calculate the covariance and correlation coefficient.
- redraw the outlier points you identified in part (2) using red color filled diamond symbol.
- add a three-line legend to your plot. The first legend line reports the covariance value; the second legend line reports the correlation coefficient value; and the third legend line indicates the red diamonds are likely outliers.

**ANSWER**
```{r, collapse=T}
attach(quakes)
cor(mag, stations)
cov(mag, stations)
plot(mag, stations)
points(mag[mag > 5.8], stations[mag > 5.8], col="red", pch=5, lwd=5)
legend("topleft", legend=c("covariance = 7.508181","correlation = 0.8511824", "outliers"), col=c("black", "black","red"), pch=c(1, 1, 5))
```

***7. Plotting Earthquakes using maps***
We can plot our earthquake records on a world map using the `maps` package. Look at the following map. Where is Fiji?

```{r, collapse=T}
library(maps)
map()
points(quakes$long, quakes$lat, pch=".", col="red")
```


*We can also zoom in and make local map of Fiji and its neighborhood area*.
```{r, collapse=T}
long.mean <- mean(quakes$long)
lat.mean <- mean(quakes$lat)
#orient <- c(lat.mean, long.mean, 0)
xlim <- c(min(quakes$long)/2, max(quakes$long)*1.5)
ylim <- c(min(quakes$lat)-10, max(quakes$lat)+10)
map(database="world", xlim=xlim, ylim=ylim, col="grey80", fill=T)
```

*Add our earthquake locations into the above plot* as red color dots. 
Used blue color filled rectangles to mark locations of the outliers identified in part (2).
```{r, collapse=T}
map(database="world", xlim=xlim, ylim=ylim, col="grey80", fill=T)
points(quakes$long, quakes$lat, pch=".", col="red")
points(quakes$long[quakes$mag > 5.8], quakes$lat[quakes$mag > 5.8], col="blue", pch=15, lwd=5)
```

