---
title: "Data Visualiation in Tidyverse"
author: "Monika Baloda"
date: 'January 10, 2023'
output: pdf_document
---

This R-markdown file uses *tidyverse* package in R to do data visualization exercise. In particular, it does the following: 

***Part I :  We use earthquake data `quakes` of R to visualize it***. 

  1. Plotting the distribution of earthquake magnitudes using tidyverse. In particular, we plot 
     i) histogram  ii) boxplot   iii) empirical cdf  iv)Q-Q plot. 
     
  2. We plot earthquakes point on top of a map layer.



***Part II : We use `mpg` dataset to use cool visualization functions of tidyverse*** 

   1. Aesthetic mapping of `color`
   
   2. Different type of smoothing curves and colored points using group identity of a variable. 
   
   3. Using _Facets_ function to add additional variable(s) to a 2D plot.
   
   4. Playing with `stat` function.
   
   5. Position adjustment options for `geom_bar()
  
   



**Load necessary packages**
```{r, collapse=T, warning=FALSE}
# install the tidyverse package first if you have not done it yet.
#install.packages("tidyverse") # you can comment out this line after you have installed `tidyverse`

library(tidyverse) # for the `ggplot2` package
```

*For rendering in PDF*
If you don't want to render (knit) the file in PDF format, you can ignore this block of code.If you face problem in rendering, please refer to debugging on https://yihui.org/tinytex/r/#debugging . 
```{r}
#tinytex::reinstall_tinytex()
```


***Main Analysis Starts From Here***

***Dataset 1: Visualization the `quakes` data set in tidyverse***
In our data visualization with Base R, we used the `quakes` data set contain the locations of 1000 seismic events of MB > 4.0. The events occurred in a cube near Fiji since 1964.

```{r}
library(datasets)
?quakes 
```

```{r, collapse=T}
class(quakes)
head(quakes, n=5) # print first 5 rows of quakes
dim(quakes)  # dimension of the table
names(quakes)  # list the variables in quakes
str(quakes)  # list the structures in quakes
glimpse(quakes) # get a glimpse of the quakes data
```

**a) Plotting the distribution of earthquake magnitudes**

Writing  `ggplot2` code to reproduce the following four subfigures in a 2-by-2 layout.  

- subfigure #1: plot a density histogram of the earthquake magnitudes, and then plot the estimated probability density curve in red color in the same plot
- subfigure #2: plot a horizontal boxplot of the earthquake magnitudes
- subfigure #3: plot the empirical cdf of the earthquake magnitudes
- subfigure #4: make a Q-Q plot to compare the observed earthquake magnitudes distribution with the Normal distribution. Add a *thick* Q-Q line in blue color.  

```{r, collpase=T, warning=F}
library(maps) 
library(datasets)
library(patchwork)

par(mfrow=c(2,2))

#subfigure-1
plot1=ggplot(data=quakes)+
geom_histogram(aes(mag), stat="density")+
xlab("Earthquake Magnitude")+
ggtitle("Histogram")+
geom_density(aes(mag), col="red")+
xlim(c(3,7))
  

#subfigure-2
plot2=ggplot(data=quakes)+
geom_boxplot(aes(mag))+
xlab("Earthquake Magnitude")+
ggtitle("Boxplot")


#subfigure-3
plot3=ggplot(data=quakes)+
stat_ecdf(aes(mag))+
xlab("Earthquake Magnitude")+
ggtitle("Emprirical CDF")

#subfigure-4
plot4=ggplot(data=quakes,mapping=aes(sample=mag))+
geom_qq()+ geom_qq_line()+
xlab("Earthquake Magnitude")+
ggtitle("Q-Q Plot")

plot1+plot2+plot3+plot4
```



**b) Earthquake location map**
Scatter plot of the earthquake locations. Use `long` as the x-axis and `lat` as the y-axis. Map `mag` to the size aesthetic and `depth` to the color aesthetic.
```{r}
ggplot(data=quakes, mapping=aes(x = long, y = lat, colour=depth, size=mag))+
geom_point()
```

**c) Plotting earthquakes point on top of a map layer**
```{r, warning=FALSE}
library(maps)
wc=map_data("world")

ggplot()+
geom_map(data=wc, map=wc, aes(long,lat, map_id=region), fill="white")+
geom_point(data=quakes, mapping=aes(x = long, y = lat, colour=depth, size=mag))
```


***Dataset 2 : Visualization of the `mpg` data set***
This data set contains fuel economy data 1999 - 2008 for 38 popular car models (source: EPA [http://fueleconomy.gov](http://fueleconomy.gov)).

```{r, collapse=T}
?mpg
dim(mpg)  # dimension of the table
names(mpg)  # list the variables in mpg
str(mpg)  # list the structures in mpg
glimpse(mpg) # get a glimpse of the mpg data
```

**a) Aesthetic mapping of `color`**

*i)* The following codes does not show points in blue color. What's gone wrong with the following code? Why are the points not blue?
```{r, collapse=T}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```
_Answer_: Points were not blue because color="blue" was written inside aes()/mapping which is not read as argument color by R, it is read as a vector c("blue") to map to an aesthetic, just like x and y variables displ and hwy. After writing color outside the mapping/aes, the R reads it as color**.

```{r, collapse=T, include=FALSE}
ggplot(data = mpg, aes(x = displ, y = hwy)) + 
  geom_point(color = "blue")
```

_Task_: Correct the code to plot blue points.
```{r, collapse=T}
ggplot(data = mpg) + 
  geom_point(aes(x = displ, y = hwy), color = "blue")
```


*(ii)* What happens if we map an aesthetic to something other than a variable name, like `aes(colour = displ < 5)`? 

_Answer_ If we map an aesthetic to something other than a variable name, like 'aes(colour= displ<5) then ggplot() function works like a temporary variable is added in the data with values equal to the result of expression. In our case, it takes values of 'TRUE' Or 'FALSE' because displ<5 is a logical variable.x and y are displ and hwy as before.

```{r, collapse=T}
ggplot(mpg, aes(x = displ, y = hwy, colour = displ < 5)) +
  geom_point()
```


**Part b) Some colorful plot using ggplot2 of tidyverse**

*i)* Smooth plots according to group identity but same color. Group is `drv` variable which is a categorical variable taking 3 values: {r,f,4}. Where `r` is for rear wheels, `f` is for front wheels, and `4` stands for all four wheels.
```{r, collapse=T}
# Enter your code here
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(group = drv), se = FALSE) +
  geom_point()
```

*ii)* Smooth plots and colors according to group identity. Group is `drv` variable which is a categorical variable taking 3 values: {r,f,4}. WHERE r IS for rear wheels,f is for front wheels, and 4 stands for all four wheels.
```{r, collapse=T}
ggplot(mpg, aes(x = displ, y = hwy, colour = drv)) +
  geom_point() +
  geom_smooth(se = FALSE)
```

*iii)* Single smooth plot but data points being colored according to group identity. Group is `drv` variable which is a categorical variable taking 3 values: {r,f,4}. WHERE r IS for rear wheels,f is for front wheels, and 4 stands for all four wheels.
```{r, collapse=T}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(se = FALSE)
```
*iv)* Diffierent smooth plots with different linetype according to group identity. Group is `drv` variable.
```{r, collapse=T}
ggplot(mpg, aes(x = displ, y = hwy)) +
  geom_point(aes(colour = drv)) +
  geom_smooth(aes(linetype = drv), se = FALSE)
```



**c) Facets:  to add additional variable(s) to a 2D plot**
There are two ways to add additional variable(s) to a 2D plot. One is using aesthetics, the other one is using facets.

*(i)* Scatter plot : x-axis is `displ` and y-axis is `hwy`. We use different colors to distinguish `drv` types.
```{r, collapse=T}
ggplot(data=mpg, mapping=aes(x = displ, y = hwy, colour=drv))+
  geom_point()
```


*(ii)* Facet `drv` into the rows. That is, makes several **rows** of subplots, one row for each `drv` type. Each subplot has `displ` mapped to the x-axis and `hwy` mapped to the y-axis. [_Note_: Use `nrow` or `ncol` to control the layout of the individual panels].

```{r, collapse=T}
 ggplot(data=mpg, aes(x=displ, y=hwy, color=drv))+
  geom_point()+
  labs(x="displacement", y="hwy")+
  scale_color_discrete(name="drv")+
  facet_wrap(~drv, nrow=1)
```



**d) `stat` functions**

Most `geom` functions and `stat` functions come in pairs that are almost always used in concert. 

- every `geom` has a default `stat`
- every `stat` has a default `geom` 

Look up the default `stat` functions for the `geom` functions listed in the following table. The variables computed by the default `stat` function (Reference: the **Computed variables** section in the R-documentation page).


| `geom` function     | default `stat` function  | variables computed by the default `stat` function  |
|-----------------    |--------------------------|-----------------------------|
| `geom_bar()`        |     count                |  Frequency                  |
| `geom_histogram()`  |     bin                  |  Frequency                  |
| `geom_density()`    |     density              |  Density, Count             |
| `geom_point()`      |     identity             |  Sum of values              |
| `geom_smooth()`     |     smooth               |  Smoothed or locally averaged  value |


_Question_: Some `geom` function has stat = "identity" as the default. What does that mean?

_Answer_ : If there are 3 teams A,B, and c with equal occurrence. Then geom_bar function will create bar chart displaying the count of occurrence in the games, which is equal. But, If we use stat="identity" with geom_bar then bar chart will be created displaying sum of points scored by the teams in each game. (Additional Note: Table formatting are sometimes tricky using R Markdown. [Table Generator](http://www.tablesgenerator.com/markdown_tables) is a handy tool if you need to make tables in the future.)


**e) Position adjustment options for `geom_bar()`**

Using two categorical variables from the `mpg` data set and to illustrate the following four position adjustment options for `geom_bar()`:

- **default**: position = "stack"
- position = "identity" will place each object exactly where it falls in the context of the graph. 
- position = "fill" works like stacking, but makes each set of stacked bars the same height. 
- position = "dodge" places overlapping objects directly beside one another. 
the bars are automatically stacked. Each colored rectangle represents a combination of cut and clarity.

*i)* position = "stack"
```{r}
ggplot(data=mpg)+
geom_bar( mapping= aes(drv) , position = "stack")

```

*ii)* position = "identity"

It will place each object exactly where it falls in the context of the graph. 
```{r}
ggplot(data=mpg)+
geom_bar( mapping= aes(drv) , position = "identity")

```

*iii)* position = "fill"

It works like stacking, but makes each set of stacked bars the same height. 
```{r}
ggplot(data=mpg)+
geom_bar( mapping= aes(drv) , position = "fill")

```

*iv)* position = "dodge"

dodge places overlapping objects directly beside one another.
```{r}
ggplot(data=mpg)+
geom_bar( mapping= aes(drv) , position = "dodge")

```


_Question_ Which position option do you like most? What conclusions can you draw from your plot?

_ANSWER_ : I liked position = "dodge" which places overlapping objects directly beside one another. Choosing right position argument is important part of making good plot. You may like other position, it is not really a right or wrong answer. 


  
  
## **Acknowledgments**
<font color="blue">
List of all the help I have received for completing this work.
1. I used rdocumentation.org website to learn more about the functions. 
2. I used geeksforgeeks.com website to get help on tidyverse paradigm of R.
</font>

