---
title: "Map visualization with `ggplot2` of Tidyverse"
author: "Monika Baloda"
output: pdf_document
---


In this R-markdown, I play with maps using tidyverse/ggplot   basic regressions.  also show some visualizations and data cleaning steps before jumping to regressions. 
Main steps done in this exercise can be put in the following points: 

  1. Plotting USA maps and state maps
  
  
  2. Plotting by manipulations
  
  
  3. Application of maps with ggplot using NYC flight destination map 


**Load necessary packages**
```{r, collapse=T,warning=FALSE}
#install.packages("maps")
#install.packages("mapproj")

library(tidyverse) # for `ggplot2`, `dplyr`, and more

library(maps) # for map visualization
#library(mapdata)
library(mapproj) 

library(datasets) # for `state` data set
library(nycflights13) # for the 2013 NYC flights data set
```

***

***Map visualization with `ggplot2`***

In this work, we will draw maps with `ggplot2`. 

The `maps` package comes with a plotting function, but, we will opt to use `ggplot2` functions (`geom_polygon` and `geom_map`) to plot the maps in the `maps` package.

Recall that `ggplot2` operates on data frames. We will use the `map_data()` function (provided by `ggplot2`), which turns a series of points along an outline into a data frame of those points.

***1) Plot the USA map***

First we load the USA map from `maps`.
```{r, collapse=T}
usa_map <- map_data("usa")
dim(usa_map)
glimpse(usa_map)
```

We use `geom_polygon()` to make a simple black map (no line color, but with a black fill).
```{r}
ggplot(data = usa_map) + 
  geom_polygon(aes(x = long, y = lat, group = group)) + 
  coord_quickmap()
```

We google the coordinates of University of California Riverside, and then use `geom_point()` to mark the location of UCR on the USA map.
In addition, change the outline/border color as well as the fill-in color of your map.

plot the USA map with white fill and black outline below
```{r, collapse=T}
usa_map= map_data("usa")

ggplot(data = usa_map) + 
  geom_polygon(aes(x = long, y = lat, group = group), fill= "white", color = "black") + 
  coord_quickmap()+
  ##point to mark the location of UCR, with coordinates of UCR
  geom_point(x = -117.2387, y = 33.9738, size= 3, colour = "red")
```

***2) Plot the states map***

In addition to `geom_polygon()`, we can use `geom_map()` to draw maps too. Basically, `geom_map()` acts as a wrapper of `geom_polygon()`. See more details in the `geom_map()` documentation @ http://ggplot2.tidyverse.org/reference/geom_map.html

Here is the example code of a states map. 
We can plot all the states, map the `fill` aesthetic to `region` and set the the lines of state borders to white color.

```{r, collapse=T}
states_map =map_data("state")
dim(states_map)
glimpse(states_map)
```

```{r, collapse=T}
ggplot() + 
  geom_map(data = states_map, map = states_map,
           aes(map_id = region, fill = region), color="white") + 
  # geom_map() doesn't work in such a way that ggplot2 knows the extend of the map values, so you always have to expand the limits by hand yourself.
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() +
  guides(fill = FALSE) # do this to leave off the color legend
```

Next, we will use the built-in `state` data sets in R to annotate our states map. In particular, `state.x77` is a two-dimensional array containing 8 variables and data from all 50 states.

```{r, collapse=T}
#?state.x77
head(state.x77)

state_data <- as.data.frame(state.x77)
state_data$State <- tolower(rownames(state_data))
state_data %>% glimpse()
```

Below is the example code from the lecture for a state population map. We first create an aesthetic mapping for `map_id` to the column `State` (state names in lower case) in the `state_data` data frame. We then call `geom_map` again and map the `fill` aesthetic to the `Population` variable in `state_data`.

```{r, collapse=T}
ggplot(data = state_data, aes(map_id = State))+ 
  geom_map(map = states_map,
           aes(fill = Population)) + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() 
```

**a)  Drawing sample states map and map the `fill` aesthetic to `Income` in the `state.x77` data set**
```{r, collapse=T}
states_map= map_data("state")

#converts the built-in dataset "state.x77" into a data frame
state_data =as.data.frame(state.x77)
state_data$State=tolower(rownames(state_data))

ggplot(data = state_data, aes(map_id = State))+ 
  geom_map(map = states_map,
           aes(fill = Income)) + 
  expand_limits(x = states_map$long, y = states_map$lat) +
  coord_quickmap() +
  labs(title = "State Income Map") +
  scale_fill_gradient(low = "white", high = "darkblue") +
  guides(fill = guide_colorbar(title = "Income"))

```

**b) Adding 50 colorful points to your map** 
We use one point to mark one state (state coordinates can be found in `state.center`). Map the color of the points to `state.region`. Map the `size` aesthetic of the points to `Population`.

```{r, collapse=T}
state_coords= data.frame(state.center)

#adds a new column "region" to the data frame "state_coords".
state_coords$region=state.region
state_coords$Population= state.x77[,"Population"]

states_map= map_data("state")

ggplot() +
  geom_polygon(data = states_map, aes(x = long, y= lat, group = group),color = "white", fill = "gray70") +
  geom_point(data = state_coords, aes(x = x, y = y, color = region, size = Population)) +
  scale_color_brewer(palette = "Dark2") +
  scale_size(range = c(1, 10)) +
  labs(title = "Population & Region of US States",
       x = "", y = "") +
 theme_void()
```
***

***3. NYC flight destination map : application of maps using ggplot***

`nycflights13::flights` data set contains all 336,776 flights that departed from New York City in 2013.

```{r, collapse=T}
#?flights # full documentation
#glimpse(flights)
```

**a) Counting the number of flights per destination**

How many unique destination airports did these NYC flights connected to? How many **non-canceled** flights per destination? Which destination had the largest number of arrival flights from NYC? Which destination had the smallest number of arrival flights from NYC?

**ANSWER**
```{r, collapse=T}
num_destinations= length(unique(nycflights13::flights$dest))
cat("Number of unique destination airports: ", num_destinations, "\n")

non_cancelled_flights= subset(nycflights13::flights, !is.na(arr_delay))
#aggregates the "arr_delay" variable
non_cancelled_by_dest= aggregate(non_cancelled_flights$arr_delay, by = list(non_cancelled_flights$dest), FUN =length)
names(non_cancelled_by_dest)= c("destination", "num_flights")
cat("Number of non-cancelled flights per destination:\n")
print(non_cancelled_by_dest)

max_dest= non_cancelled_by_dest[which.max(non_cancelled_by_dest$num_flights),"destination"]
#cat function to print a message to the console
cat("Destination with the largest number of arrival flights: ", max_dest, "\n")

min_dest=non_cancelled_by_dest[which.min(non_cancelled_by_dest$num_flights), "destination"]
cat("Destination with the smallest number of arrival flights: ", min_dest, "\n")
```

**b) Marking all destination airports on a states map**
Find out the coordinates of the destination airports from `nycflights13::airports`. Draw each destination airport as a point on a states map, and map a point aesthetic to the number of **non-canceled** flights flew to that destination from NYC in 2013.

```{r, eval=F}
#?airports
airports %>% glimpse()
left_join(per_dest, airports, by=c("dest"="faa"))
```

```{r, collapse=T}
library(ggplot2)
library(nycflights13)
library(magrittr)
library(dplyr)

per_dest= flights %>%
  filter(!is.na(dep_delay)) %>% # remove cancelled flights
  group_by(dest) %>% #groups the filtered data by the dest column
  summarize(n = n()) # count the non-cancelled flights/destination

 #performs a left join between the per_dest & airports dataframe
df= left_join(per_dest, airports, by = c("dest" = "faa"))

ggplot(df, aes(x = lon, y =lat, size = n)) +
  borders("state", colour = "grey50", fill = NA) +
  geom_point() +
  scale_size_continuous(range = c(1, 10))
```
_Question_ Which destination airports have missing values? 

_ANSWER_
```{r, collapse=T}
#here dep_delay column is not NA (i.e., missing values).
per_dest = flights %>% 
  filter(!is.na(dep_delay)) %>% 
  group_by(dest) %>% 
  summarize(n = n()) %>% 
  arrange(desc(n))

# to find the airports with missing
missing_airports = flights %>% 
  filter(is.na(dep_delay)) %>% 
  distinct(dest)

# Print the list of airports with missing values
#missing_airports that shows the unique destinations 
missing_airports
```