---
title: "Basic Questions in R"
author: "Monika Baloda"
output: pdf_document
---


***This document takes data from Prof. Flegal of Statistics department, UCR's course Statistical Computing : STAT-206 homework problems***

```{r}
#tinytex::install_tinytex()  #to knit the R-markdown in PDF format
```

1. The data set at [http://www.faculty.ucr.edu/~jflegal/206/rnf6080.dat] records hourly rainfall at a certain location in Canada, every day from 1960 to 1980.
    
    1a). First, we need to load the data set into R using the command `read.table()`. Use the help function to learn what arguments this function takes. Once you have the necessary input, load the data set into R and make it a data frame called `rain.df`.

***Answer of 1a)***    
```{r}
rain.df=read.table("http://www.faculty.ucr.edu/~jflegal/206/rnf6080.dat")

```
    
  1b). How many rows and columns does `rain.df` have? (If there are not 5070 rows and 27 columns, something is wrong; check the previous part to see what might have gone wrong in the previous part.)
  
***Answer 1b)*** There are 5070 number of rows and 27 number of columns. We find out this by dim function which gives us number of dimensions.
```{r}
dim(rain.df)

```
    1c). What are the names of the columns of `rain.df`?
***Answer 1c)***: The variables are just named as V1, V2, V3,....,V27. 

```{r}
names(rain.df)
```
    1d). What is the value of row 5, column 7 of `rain.df`?
***Answer 1d)*** The value is 0. We access it by simply supplying the row and column dimension numbers as follows.
```{r}
      rain.df[5,7]
```
    1e) Display the second row of `rain.df` in its entirety.
***Answer 1e)*** We print the second row by leaving the column argument empty which means print all the columns (default setting).
```{r}
print(rain.df[2,])    
```
    1f). Explain what this command does:
***Answer 1f)***  We note that the command renames the first three columns to be 'year', 'month', 'day', and rest of the 24 variables are named         as numerical numbers. The remaining variables shows daily rainfall. 
```{r}
 names(rain.df) <- c("year","month","day",seq(0,23))
#head(rain.df)
```
  1g). Create a new column in the data frame called `daily`, which is the sum of the rightmost 24 columns. 
    With this column, create a histogram of the values in this column, which are supposed to be daily rainfall values. 
    What is wrong with this picture?
    
***Answer 1g)***: Negative value of rainfall does not make sense. This histogram shows us some negative values of rainfall. 
```{r}
rain.df$daily= apply(rain.df[,c(4:27)], 1, sum)
hist(rain.df$daily, xlab="Daily rainfall", main="Histogram of daily rainfall")
```
    1h) Create a new data frame `rain.df.fixed` that takes the original and fixes it for the apparent flaw you have discovered. Having done this,          produce a new histogram with the corrected data and explain why this is more reasonable.
***Answer 1h)***: The following codes fixes the issue in 1g). 
     
```{r}
      rain.df.fixed=rain.df
      rain.df.fixed$daily=ifelse(rain.df.fixed$daily<0, rain.df.fixed$daily*0, rain.df.fixed$daily)
      hist(rain.df.fixed$daily, xlab="Daily rainfall", main="Histogram of daily rainfall : Fixed")
```

2. Syntax and class-typing.

   2a. For each of the following commands, either explain why they should be errors, or explain the non-erroneous result.
    
***Answer 2a)***: vector1 is a vector of 4 objects stored in character format. It appears that these are numbers but R doesn't know that they are      numbers. R treats them as character type objects. That is why when sorting it, R places 12 before 5 because the first 1 comes before 5, it            doesn't read it as 12 numeric. Further, it cannot sum these numbers because the numbers are not stored as numbers. 
```{r, eval=FALSE}
      vector1 <- c("5", "12", "7", "32")
      max(vector1)
      sort(vector1)
      sum(vector1)
```

  2b. For the next series of commands, either explain their results, or why they should produce errors.

```{r, eval=FALSE}
vector2 <- c("5",7,12)
vector2[2] + vector2[3]

dataframe3 <- data.frame(z1="5",z2=7,z3=12)
dataframe3[1,2] + dataframe3[1,3]

list4 <- list(z1="6", z2=42, z3="49", z4=126)
list4[[2]]+list4[[4]]
list4[2]+list4[4]

class(vector2)
class(dataframe3[1,2])
  
```
***Answer 2b)***:
     1. vector2[2] + vector2[3] is not possible because the class of vector2 is character and not numeric, therefore the plus sign cannot do the           addition arithmetic. Below we show that vector2 is of character type. 
    2. Dataframes are allowed to store different classes of objects together. dataframe3[1,2] returns numerical(double) type and dataframe3[1,3] also     returns numerical type objects, therefore their addition is possible.  
    3. list4[2]+list4[4] doesn't work because list4[2] is a list type object on which we cannot do addition operation. But list4[[2]] returns numeric     (double) type, therefore it is possible to do arithmetic operation like +. 

    


3. Working with functions and operators.

  3a. The colon operator will create a sequence of integers in order. It is a special case of the function `seq()` which you saw earlier in this assignment. Using the help command `help(seq)` to learn about the function, design an expression that will give you the sequence of numbers from 1 to 10000 in increments of 372. Design another that will give you a sequence between 1 and 10000 that is exactly 50 numbers in length.
   
***Answer 3a)***: The following block of codes generated the desired sequences asked in the question. 
```{r}
    vec1=c(1:10000)
    vec2=seq(from=1,to=10000,by=372)
    vec3=seq(from=1,to=10000,length.out= 50)
```

  3b. The function `rep()` repeats a vector some number of times. Explain the difference between `rep(1:3, times=3) and rep(1:3, each=3).
    
***Answer 3b)***: The command rep(1:3, times=3) generates a series: 1 2 3 1 2 3 1 2 3 while the latter command rep(1:3, each=3) generates the     series 1 1 1 2 2 2 3 3 3. Meaning that first command goes on to complete the sequence and then repreat it while the second command first repeat and then execute the sequence completion. 


